PyUltraLight: A Pseudo-Spectral Solver for Ultralight Dark Matter Dynamics

### 2018-10-01 draft.tex
Latest revisions according to RE's comments

\usepackage[T1]{fontenc} % if needed
\usepackage{eucal}
\usepackage{xspace}
\usepackage{color}
\usepackage{url}
\newcommand{\re}[1]{\textcolor{blue}{[{\bf RE}: #1]}}
\newcommand{\ek}[1]{\textcolor{red}{[{\bf EK}: #1]}}

@@

where we subtract the average density from the right hand side of the Poisson equation. The form of Equation \ref{eq:poisson_periodic} is a consequence of Gauss' law and the fact that the surface integral of the gradient of the field is identically zero when periodic boundary conditions are imposed.

\ek{I've given slightly more info here, think this is sufficient to understand the form of 2.8} \re{I think this is mostly there, but we should say which surface integral and may cite some of the literature here}

@@

The default timestep in  \PyUltraLight  is fixed with reference to the fluid interpretation of the Schr{\"o}dinger-Poisson system \cite{Hui:2016ltb}. The fluid interpretation is often used to recast the Schr{\"o}dinger-Poisson system in the form of the Madelung equations \cite{Suarez:2011yf}, which a hydrodynamical representation of the system. The first step is to define
\begin{equation}\label{eq:madelung}
    \psi\equiv\sqrt{\rho}e^{i\theta}, \quad \vec{v}\equiv\boldsymbol{\nabla}\theta.
\end{equation}
and to treat $\vec{v}$ as a fluid velocity.  From this perspective, if the phase difference between two adjacent grid points exceeds $\pi$ the fluid will appear to move ``backwards'' across the grid. We thus set the default timestep, $\Delta t$, so that fluid travelling at this maximum velocity traverses one grid space, $\Delta x$, per timestep, or
\begin{equation}\label{eq:timestep}
    \Delta t = \frac{(\Delta x)^2}{\pi}.
\end{equation}
\re{Tweaked a little; changed phase change to phase difference; the latter sounds more explicitly spatial}
This is a choice, rather than a strict constraint on $\Delta t$. However, if the ``fluid'' approaches velocities where the phase appears to switch direction, the configuration is approaching the point where the simulation grid is too coarse to fully resolve the dynamics. Hence, a timestep much smaller than this value may offer little practical advantage. However, in some cases the breakdown may occur in regions of the simulation volume that are of little physical interest, and the user is free to choose a larger timestep  via the 'step\textunderscore factor' parameter in the Jupyter notebook.

@@

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 1cm 0 2cm,clip]{combined2}
  \caption{Top: Deviation of the phase of $\psi$ compared to the highest resolution result ($320^3$). Field values are arranged in order of increasing magnitude from left to right. A slight improvement in phase convergence can be seen for higher density regions to the right. Bottom: Improving convergence of $\vert\psi\vert$ with increased spatial resolution for the simulation shown in Figure \ref{fig:disruption}}
  \label{fig:spatial}
\end{figure}

\begin{figure} 
  \includegraphics[width=1.\textwidth,trim=0 1cm 0 2cm,clip]{combined}
  \caption{Top: Phase deviation of $\psi$, relative to solution with timestep 0.1 times default, sorted by the density. There is excellent agreement with the default timestep, and reasonable convergence at steps up to $10$ times the default, with better accuracy in high density regions. Bottom: Difference in magnitude of $\psi$, relative to the  solution with timestep 0.1 times default. Again we see good convergence with the default timestep, and tolerable agreement in high density regions when the step is a factor of $10$ or less than default. \re{Do we want to the put the ``key'' at the bottom not the middle; also I think the $x$-axis label could be clearer, or dropped.}}
  \label{fig:temporal}
\end{figure}

Figures \ref{fig:comparison} and \ref{fig:comparison2} show profiles of the density through the simulation volume, as a function of spatial resolution and timestep.  The figures represent the density profiles down the axis of symmetry of the initial configuration (vertical axis in Figure \ref{fig:disruption}) after approximately half a revolution around the central potential, or t=0.28 code units -- slightly after the final frame in Figure \ref{fig:disruption} - when the solitons have become distorted due to tidal forces, but are not yet completely disrupted. Figure \ref{fig:comparison} shows results obtained when the timestep is varied from 0.1 to 50 times the default value. The results with the default and the shorter timestep are virtually indistinguishable, and results are still reasonably accurate at 10 times the default timestep, with small deviations at high densities. However, the results are significantly distorted at 50 times the default timestep. Figure \ref{fig:comparison2} shows results at resolutions from $64^3$ to $320^3$. The lowest resolution performs poorly, but there is good convergence  at resolutions of $192^3$ and above. \re{minor edits; we could do the two figures as a single image}

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 0.8cm 0 1.2cm,clip]{comparison}
  \caption{Effect of increasing the timestep on the density profile at half a revolution.}
  \label{fig:comparison}
\end{figure}

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 0.8cm 0 1.2cm,clip]{comparison2}
  \caption{Effect of decreasing the spatial resolution on the density profile at half a revolution.}
  \label{fig:comparison2}
\end{figure}

>>>>

\usepackage[T1]{fontenc} % if needed
\usepackage{eucal}
\usepackage{xspace}
\usepackage{color}
\usepackage{url}
\usepackage{afterpage}
\newcommand{\re}[1]{\textcolor{blue}{[{\bf RE}: #1]}}
\newcommand{\ek}[1]{\textcolor{red}{[{\bf EK}: #1]}}

@@

where we subtract the average density from the right hand side of the Poisson equation. The form of Equation \ref{eq:poisson_periodic} is a consequence of Gauss' law and the fact that the surface integral of the gradient of the field around the perimeter of the simulation grid is identically zero when periodic boundary conditions are imposed \cite{Dabo:2008}.

\re{I think this is mostly there, but we should say which surface integral and may cite some of the literature here}\ek{How's this? It seems to be difficult to find many references which explain the subtraction of the average in detail, a lot of people just seem to use it without explanation.}

@@

The default timestep in  \PyUltraLight  is fixed with reference to the fluid interpretation of the Schr{\"o}dinger-Poisson system \cite{Hui:2016ltb}. The fluid interpretation is often used to recast the Schr{\"o}dinger-Poisson system in the form of the Madelung equations \cite{Suarez:2011yf}, which a hydrodynamical representation of the system. The first step is to define
\begin{equation}\label{eq:madelung}
    \psi\equiv\sqrt{\rho}e^{i\theta}, \quad \vec{v}\equiv\boldsymbol{\nabla}\theta.
\end{equation}
and to treat $\vec{v}$ as a fluid velocity.  From this perspective, if the phase difference between two adjacent grid points exceeds $\pi$ the fluid will appear to move ``backwards'' across the grid. We thus set the default timestep, $\Delta t$, so that fluid travelling at this maximum velocity traverses one grid space, $\Delta x$, per timestep, or
\begin{equation}\label{eq:timestep}
    \Delta t = \frac{(\Delta x)^2}{\pi}.
\end{equation}
This is a choice, rather than a strict constraint on $\Delta t$. However, if the ``fluid'' approaches velocities where the phase appears to switch direction, the configuration is approaching the point where the simulation grid is too coarse to fully resolve the dynamics. Hence, a timestep much smaller than this value may offer little practical advantage. However, in some cases the breakdown may occur in regions of the simulation volume that are of little physical interest, and the user is free to choose a larger timestep via the 'step\textunderscore factor' parameter in the Jupyter notebook.

@@

Figure \ref{fig:comparison} shows profiles of the density through the simulation volume, as a function of spatial resolution and timestep.  Each plot represents the density profile down the axis of symmetry of the initial configuration (vertical axis in Figure \ref{fig:disruption}) after approximately half a revolution around the central potential, or t=0.28 code units -- slightly after the final frame in Figure \ref{fig:disruption} - when the solitons have become distorted due to tidal forces, but are not yet completely disrupted. We see that as the timestep is varied from 0.1 to 50 times the default value, the results with the default and the shorter timestep are virtually indistinguishable, and results are still reasonably accurate at 10 times the default timestep, with small deviations at high densities. However, the results are significantly distorted at 50 times the default timestep. We also see that as the spatial resolution is decreased from $320^3$ to $64^3$, the lowest resolution performs poorly, but there is good convergence  at resolutions of $192^3$ and above. \re{minor edits; we could do the two figures as a single image}\ek{How's this?}

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 1cm 0 2cm,clip]{combined2}
  \caption{Top: Deviation of the phase of $\psi$ compared to the highest resolution result ($320^3$). Field values are arranged in order of increasing magnitude from left to right. A slight improvement in phase convergence can be seen for higher density regions to the right. Bottom: Improving convergence of $\vert\psi\vert$ with increased spatial resolution for the simulation shown in Figure \ref{fig:disruption}}
  \label{fig:spatial}
\end{figure}
\clearpage

\begin{figure} 
  \includegraphics[width=1.\textwidth,trim=0 1cm 0 2cm,clip]{combined}
  \caption{Top: Phase deviation of $\psi$, relative to solution with timestep 0.1 times default, sorted by the density. There is excellent agreement with the default timestep, and reasonable convergence at steps up to $10$ times the default, with better accuracy in high density regions. Bottom: Difference in magnitude of $\psi$, relative to the  solution with timestep 0.1 times default. Again we see good convergence with the default timestep, and tolerable agreement in high density regions when the step is a factor of $10$ or less than default. \re{Do we want to the put the ``key'' at the bottom not the middle; also I think the $x$-axis label could be clearer, or dropped.}}\ek{How's this?}
  \label{fig:temporal}
\end{figure}
\clearpage

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 1.2cm 0 2.0cm,clip]{comparison}
  \caption{Top: Effect of decreasing the spatial resolution on the density profile at half a revolution. Bottom: Effect of increasing the timestep on the density profile at half a revolution.}
  \label{fig:comparison}
\end{figure}
\clearpage
